file name: PA1_template.Rmd
```{r echo=FALSE, setup} 
    rm(list=ls()) # Clear out workspace
    knitr::opts_chunk$set(echo=TRUE, warning=TRUE)
```

# Peer Assignment 1

This assignment makes use of data from a personal activity monitoring device. 
This device collects data at 5 minute intervals through out the day. The data 
consists of two months of data from an anonymous individual collected during the 
months of October and November, 2012 and include the number of steps taken in 
5 minute intervals each day.

This assignment will examine how the number of steps taken varies with time of 
day and day of week


**First load libraries and data**

```{r load file}

    library(lubridate)

    file.name <- "activity.zip"

    # using http instead of https:
    URL <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
    download.file(URL, destfile = file.name)
    unzip(file.name)

    activity <- read.csv("activity.csv")

```

    File `r print(file.name)` downloaded on `r date()`

    File contains variables: `r names(activity)`

**Now calculate mean and median steps per day**

```{r mean calc}

    first.day     <- ymd(as.character(activity$date[1]))  # first day of test
    date_sv       <- ymd(as.character(activity$date))     # full date of each measurement
    # Calculate a relative day, e.g., 0,1,2....
    activity$date <- 
        round(as.numeric(as.duration(
            interval(first.day, ymd(as.character(activity$date)))))/86400)

    # Rename table variable
    names(activity)[2] <- "day"

    loc <- !is.na(activity$steps)  # locations of valid data

    # Calculate total, mean, and median steps per day
     total.stepsperday  <- tapply(activity$steps[loc],activity$day[loc],sum)
      mean.stepsperday  <- mean(total.stepsperday)
    median.stepsperday  <- median(total.stepsperday)
```

    Reformatted file contains variables: `r names(activity)`  

  **Mean steps per day is: `r format(mean.stepsperday, scientific=FALSE)`**    
**Median steps per day is: `r median.stepsperday`**    

```{r Hist}
    # Form histogram
    hist(total.stepsperday,breaks=11,xlim=c(0,25000),
         main="Histogram of Total Steps per Day",
         xlab="Total Steps per Day")

```



```{r steps per int}

    # Calculate mean steps per interval
    mean.stepsperinterval <- tapply(activity$steps[loc],as.factor(activity$interval[loc]),mean)
    plot(mean.stepsperinterval, type="l", 
         ylab="Steps per Interval", 
         xlab="5 Minute Interval Number, within a day")
    title(main="Average Daily Activity Pattern")
```

```{r max}

    # Find max
    max.stepsperinterval <- max(mean.stepsperinterval)
    mloc <- which.max(mean.stepsperinterval)
```

**Max Steps per interval is `r format(max.stepsperinterval, digits=4)`**    
**Max occurs at interval number `r mloc` or minute `r activity$interval[mloc]` of the day**  

**Now correct for missing data:**

```{r miss}

        nmiss <- sum(is.na(activity$step))  # number of intervals with missing data

        activity_nm <- activity             # initialize new table

        # Replace missing data with mean for that interval

        # Load means into places where data is missing
        for (i in 1:17568){
            if (is.na(activity_nm$steps[i])) {
                j <- i%%288 + 1
                activity_nm$steps[i] <- mean.stepsperinterval[j]
            }
        }

        # Calculate new mean after data correction
        mean.stepsperinterval2 <- tapply(activity_nm$steps, as.factor(activity_nm$interval),mean)

        # Calculate difference in pre/post correction means
        diffs <- mean.stepsperinterval2-mean.stepsperinterval

        # Find largest difference
        max_diff <- max(abs(diffs))

        # Calculate corrected statistics
         total.stepsperday2  <- tapply(activity_nm$steps,activity_nm$day,sum)
          mean.stepsperday2  <- mean(total.stepsperday2)
        median.stepsperday2  <- median(total.stepsperday2)

```

A total of `r nmiss` intervals are missing data  

**Max difference in steps per day due to replacing NAs with mean of 5 min values is `r format(max_diff, digits=3)`**    

```{r hist2}
        # Form histogram
        hist(total.stepsperday2,breaks=11,xlim=c(0,25000),
             main = paste("Histogram of Total Steps per Day, after NA removal"),
             xlab="Total Steps per Day")
```

**Corrected statistics:**  
Mean Steps per Day `r format(mean.stepsperday2, scientific=FALSE)`  
Median Steps per Day `r format(median.stepsperday2, scientific=FALSE)`  

**The mean is unchanged, but now the median equals the mean**

```{r final figs, fig.height=8}

wkday <- weekdays(date_sv)

loc_wend <- (wkday=="Saturday") | (wkday=="Sunday")

daytype <- rep("weekday",17568)
daytype[loc_wend] <- rep("weekend",sum(loc_wend))
                        
activity_nm$daytype <- as.factor(daytype)

mean.weekday  <- tapply(activity_nm$steps[!loc_wend],as.factor(activity_nm$interval[!loc_wend]),mean)
mean.weekend  <- tapply(activity_nm$steps[ loc_wend],as.factor(activity_nm$interval[ loc_wend]),mean)
  
par(mfrow=c(2,1))

plot(mean.weekend,type="l",main="Average Weekend Steps/Interval",xlab="Interval",ylab="Average Steps per Interval")
plot(mean.weekday,type="l",main="Average Weekday Steps/Interval",xlab="Interval",ylab="Average Steps per Interval")

```